Probabilistic Forward Modeling of Galaxy Catalogs with Normalizing Flows

### 2023-11-22 src/tex/ms.tex
Lots of manuscript changes.

% Title
\title{Statistical Modeling of Galaxy Catalogues with Normalizing Flows}

@@

\begin{abstract}
    Normalizing flows are powerful tools for learning high-dimensional probability distributions from samples thereof, and have many applications in astronomy, including posterior estimation and forward modeling.
    We introduce PZFlow, a Python package for statistical modeling of tabular data using normalizing flows.
    We use PZFlow to model photometric galaxy catalogs including redshifts, photometry, size, and shape, and generate a synthetic catalog.
    In this catalog, each galaxy has a \emph{true} redshift posterior.
    We discuss the importance of these true posteriors for the comprehensive evaluation of redshift posteriors produced by photometric redshift (photo-z) estimators.
    We also demonstrate the use of an ensemble of normalizing flows for density estimation, applied to photo-z estimation.
    While we focus on photo-z estimation and validation, we emphasize that these methods are applicable to any galaxy properties, and any other tabular data.
\end{abstract}

@@

Astronomical data represent realizations of complex probability distributions.
A common goal in research is to infer the underlying distribution from a limited set of noisy data, and use this distribution to estimate posterior distributions over galaxy and population parameters.
An important example is photometric redshift (photo-z) estimation, where galaxy redshift posteriors are estimated from galaxy photometry, using a model informed by a training set of spectroscopic galaxy data \citep{newman2022}.
These redshift posteriors are then used to estimate posterior distributions for cosmological parameters \citep{descSRD}.

Knowledge of the probability distributions underlying our data is also valuable for simulation and forward modeling astronomical data sets.
Simulating data sets with realistic statistical properties enables methodological development and the calibration of systematic uncertainties.
Forward modeling also facilitates data augmentation (e.g. \citealt{lokken2022}) and the creation of large data challenges (e.g. \citealt{kessler2019, dc2, cosmodc2}), which are becoming more prevalent in the big data era of astronomy.

Inferring the underlying probability distribution, or likelihood, from a high-dimensional data set is a difficult problem.
Many machine learning tools have been developed for this task, including Generative Adversarial Networks (GANs; \citealt{goodfellow2014}) and Variational Autoencoders (VAEs; \citealt{kingma2014}).
Both are neural networks that excel at forward modeling complex data sets, but neither allow exact likelihood calculation for the simulated data.
This means that questions such as ``under my model, what is the posterior distribution for redshift given the simulated photometry'' have only approximate answers.

Normalizing flows, on the other hand, are a deep learning tool that excel at forward modeling, while also allowing exact, analytic likelihood calculation.
In other words, they allow you to analytically calculate likelihoods and posteriors with respect to the distribution from which simulated data is drawn.
Normalizing flows operate by learning an invertible transformation of the data distribution into a simpler, tractable distribution, known as the latent distribution.
A common choice for latent distribution is a normal distribution, hence the name \emph{normalizing} flow.
This allows sampling and likelihood calculation to be performed within the context of the simple latent distribution, with the normalizing flow acting as a translator between the latent samples and likelihoods, and their corresponding values in the context of the complicated data distribution.

Normalizing flows have gained popularity as tools for efficient and flexible sampling for parameter inference (e.g., \citealt{dai2022,dacunha2022,hassan2022}).
In this paper, we focus on photo-z's and forward modeling photometric galaxy catalogs.
Since normalizing flows allow exact, analytic probability calculation, the properties of objects in these catalogs have \emph{true} posteriors.
This means that questions such as ``under my model, what is the posterior distribution for redshift given the simulated photometry'' have exact answers.
Catalogs with true posteriors are useful for the testing and validation of analysis pipelines that estimate posteriors, such as the photo-z estimators used in much of astrophysics and cosmology.
Previous evaluations of these estimators have focused on comparing point estimates to true values (e.g., \citealt{hildebrandt2010,sanchez2014,graham2018}), or evaluating ensembles of posteriors \citePZp.
Normalizing flow catalogs with true posteriors open up a new, more comprehensive avenue for evaluation of these estimators by enabling direct posterior-to-posterior comparison.

To facilitate the statistical modeling of galaxy catalogs and other astronomical data sets, we have developed PZFlow, a normalizing flow package for Python.
With relatively little tuning required by the user, PZFlow can provide a generative model for any tabular data, including continuous and discrete variables, and variables with Euclidean or periodic topology.
In addition to generative modeling, PZFlow can calculate posteriors over any columns in your data set, and can convolve errors and marginalize over missing columns while training the model or calculating posteriors for samples.

In this paper, we provide the background on normalizing flows (Section \ref{sec:nf}) required to understand PZFlow (Section \ref{sec:pzflow}).
We then use PZFlow to simulate a galaxy catalog (Section \ref{sec:galaxy-catalog}), where each object has photometry, size, ellipticity, redshift, and a true redshift posterior.
We also demonstrate using PZFlow as a density estimator, via the example of photo-z estimation (Section \ref{sec:photo-z}).
We conclude in Section \ref{sec:conclusion}.

@@

If we further assume that $f$ is invertible, we can sample from the data distribution by applying $f^{-1}$ to samples from the latent distribution\footnote{Here, $~$ means ``is drawn from.''}:
\begin{align}
    x = f^{-1}(u) \quad \text{where} \quad u \sim \pu.
\end{align}

@@

A bijection is an invertible map between two sets.
In a normalizing flow, the bijection maps the data distribution onto the latent distribution for likelihood calculation, and the inverse of the bijection maps samples from the latent distribution back to the data distribution.
The bijection of a normalizing flow must be powerful enough to model complex relationships in data, while remaining invertible and simultaneously possessing an efficiently computable Jacobian determinant.
This latter constraint is the primary difficulty in designing a normalizing flow.
The most popular strategy for achieving these requirements is to exploit the fact that a composition of bijections is also bijective.
By chaining together multiple less-expressive bijections whose Jacobians are efficiently computable, a composite bijections can be constructed that meets our requirements:
\begin{align}
    f &= \dots \circ f_3 \circ f_2 \circ f_1.
\end{align}
The overall Jacobian determinant can be efficiently calculated using the chain rule.

@@

In general, as long as sampling and density evaluation are tractable, one can use any parameterization of the latent distribution that matches some desired structure in the data and learn the distribution parameters during training.
We give this generalization the name \emph{latent-adaptive flows} (LAFs; inspired by the Tail Adaptive Flows of \citealt{jaini2020}).
Our experiments indicate that learnable latent distributions can improve training loss, but require more care in training.

@@

PZFlow is a Python package for building normalizing flows, with a focus on easy high-performance modeling of high-dimensional tabular data.
Data is handled in Pandas DataFrames \citep{pandas}, while the normalizing flows are implemented in Jax \citep{jax}, which allows for efficient, parallelizable, GPU-enabled calculations for very large data sets.
The code is easily installable from the Python Package Index\footnote{\url{https://pypi.org/project/pzflow/}} (PyPI) and is hosted on Github,\footnote{\url{https://github.com/jfcrenshaw/pzflow}}.
The documentation\footnote{\url{https://jfcrenshaw.github.io/pzflow/}} includes tutorial notebooks demonstrating the features mentioned in this paper on different example problems.

The rest of this paper will demonstrate using PZFlow for the statistical modeling of galaxy catalogs.
Section \ref{sec:galaxy-catalog} uses PZFlow to forward model a galaxy catalog, including photometry, spectroscopic redshifts (spec-z's), \emph{true} photo-z posteriors, ellipticities, and sizes.
Section \ref{sec:photo-z} uses PZFlow for photo-z estimation, demonstrating the power of PZFlow as a density estimator, including numerous useful features for photo-z estimation.

In addition to the examples in this paper, PZFlow has already being used in various other projects:
\begin{itemize}
    \item \citet{malz2021} used PZFlow to build a metric for observing strategy optimization based on information theory;
    \item \citet{stylianou2022} used PZFlow to forward model galaxy data with true redshift posteriors in order to evaluate the impact of survey incompleteness and spec-z errors on photo-z estimation;
    \item \citet{lokken2022} used PZFlow to smooth high-redshift artifacts in simulations of host galaxies for supernovae and other transients.
\end{itemize}

@@

In this section, we use PZFlow to forward model a photometric galaxy catalog for the Vera Rubin Observatory's Legacy Survey of Space and Time (LSST; \citealt{ivezic2019}).
The advantage of using a catalog generated from a normalizing flow is that we have direct access to the exact distribution from which the data is drawn, enabling us to calculate true values for derived statistical products, such as the \emph{true} photo-z redshift posterior for each galaxy.

@@

\begin{figure*}[t]
    \script{forward_model/plot_main_galaxy_corner.py}
    \begin{centering}
        \includegraphics{figures/main_galaxy_corner.pdf}
        \caption{
            Distribution of true redshifts and noiseless photometry from the CosmoDC2 test set, compared to a sample drawn from the distribution learned by PZFlow.
            The close overlap of every pair-wise distribution demonstrates that PZFlow has learned the distribution in CosmoDC2 with high fidelity.
        }
        \label{fig:main-corner}
    \end{centering}
\end{figure*}

\begin{figure*}[t]
    \script{forward_model/plot_smooth_color_distribution.py}
    \begin{centering}
        \includegraphics{figures/smooth_color_distribution.pdf}
        \caption{
            Comparing the noiseless $r-i$ vs redshift distribution for galaxy samples from CosmoDC2 (left) and from the normalizing flow (right).
            The high-redshift galaxies in CosmoDC2 lie along discrete tracks in color space due to the discrete number of galaxy SED templates used in the simulation.
            PZFlow smooths over these discrete tracks, resulting in a color distribution that is smooth to high redshifts.
        }
        \label{fig:smooth-color-dist}
    \end{centering}
\end{figure*}

To create a generative model of galaxy redshifts and photometry, we use the true redshifts and $ugrizy$ magnitudes from the CosmoDC2 simulation \citep{dc2, cosmodc2} of the LSST Dark Energy Science Collaboration (DESC).
We selected all galaxies from CosmoDC2 with at most one band with a signal-to-noise ratio (SNR) less than 10, using the forecasted photometric errors from LSST year 10.
Of these, we randomly selected $10^6$ galaxies and split them intro training and test sets consisting of 80\% and 20\% of the galaxies, respectively.

@@

After training the flow (see Appendix \ref{app:training-details}), we assess the results by drawing $10^4$ galaxies from the trained flow, and plotting their distribution against  $10^4$ galaxies from the test set (Figure \ref{fig:main-corner}).
We see the normalizing flow has done an excellent job of reproducing the distribution of galaxies in CosmoDC2, without any unusual artifacts or outliers.
In addition, Figure \ref{fig:smooth-color-dist} compares the distribution of galaxy $r-i$ vs redshift.
The CosmoDC2 simulation is known to exhibit discrete tracks in this space at high redshift, due to the discrete number of SED templates used during simulation.
These tracks are visible in the left panel.
The right panel shows that PZFlow smooths over this discreteness, resulting in a color distribution that is smooth up to high redshifts.

We note that these results were obtained without any extensive hyperparameter search, and that very similar (slightly worse results) are obtained without the \texttt{ColorTransform} bijection, demonstrating the flexibility of the method to adapt to unseen data sets.

With this normalizing flow, we have an efficient CosmoDC2 emulator that produces a smooth distribution of realistic galaxies up to high-redshifts.
We use this emulator to generate a catalog with $10^4$ galaxies.
We add photometric errors using the LSST error model of our PhotErr package (see Appendix \ref{app:error-model}).
Importantly, since we have access to the probability distribution from which the galaxies were generated, we can calculate \emph{true} redshift posteriors for each galaxy.
This is the subject of the next section.

@@

\begin{figure}[t]
    \script{forward_model/plot_posteriors.py}
    \begin{centering}
        \includegraphics{figures/posteriors.pdf}
        \caption{
            True redshift posteriors for a galaxy representing different amounts of information.
            The black posterior is calculated with the true magnitudes;
            the blue posterior is calculated after adding photometric errors;
            the orange posterior is calculated after adding photometric errors, but with the errors convolved during posterior calculation;
            the green posteriors is the same as the orange, except with the $u$ band marginalized over.
        }
        \label{fig:posteriors}
    \end{centering}
\end{figure}

@@

Since we have direct access to the probability distribution from which the photometry and redshifts are drawn, we can analytically calculate the true redshift posterior for each galaxy: $p(z|m)$ where $m$ is the vector of galaxy magnitudes.
We note that this is not an estimate, like what would be returned by a photo-z estimator, but rather the truth, obtained from the model that generated the photometry and redshifts in the first place.

In addition to calculating the true posterior, $p(z|m)$, we can calculate a true posterior that is consistent with the photometric errors:
\begin{align}
    p(z|m, \sigma_m) = \int p(z|\hat{m}) p(\hat{m}|m, \sigma_m) d\hat{m},
    \label{eq:err-conv}
\end{align}
where $\sigma_m$ is the vector of photometric errors returned by the error model, and $\hat{m}$ are possible noisy observations of the true magnitudes, $m$.
We can evaluate this integral numerically by sampling $\hat{m} \sim p(\hat{m}|m, \sigma_m)$, evaluating $p(z, \hat{m})$ on a redshift grid, and normalizing to obtain $p(z|\hat{m})$.
Averaging $p(z|\hat{m})$ over the samples $\hat{m}$ yields $p(z|m, \sigma_m)$.
This is the ``best case scenario'' redshift posterior that photo-z estimators hope to estimate.

We can also marginalize over any missing bands, $n$:
\begin{align}
    p(z|\hat{m}) = \frac{1}{p(\hat{m})} \int p(z, n, \hat{m}) dn,
\end{align}
which can be calculated by evaluating $p(z, n, \hat{m})$ on a grid of $z$ and $n$, summing over $n$ to yield $p(z, \hat{m})$, and normalizing with respect to redshift to yield $p(z|\hat{m})$.
You can once again average over $\hat{m}$ samples to convolve the photometric errors.
You may wish to marginalize over all values of $n$ if the galaxy was not observed in that band.
This might occur, for example, if you include photometry from the Euclid Space Telescope \citep{euclid}, which will not have complete coverage of the LSST catalog.
You may also wish to marginalize over all values beyond a limiting magnitude if the galaxy was observed, but not detected in that band.
This might occur, for example, in the low wavelength bands of Lyman dropout galaxies observed by LSST.

To visualize the impacts of error convolution and band marginalization, Figure \ref{fig:posteriors} shows a number of redshift posteriors for a single example galaxy.
The black posterior is calculated using the true galaxy magnitudes, while the blue posteriors is calculated after adding photometric errors.
Calculating the posterior using the noisy photometry results in a biased posterior.
The orange posterior has had the photometric errors convolved as in Equation \ref{eq:err-conv}.
Convolving the errors broadens the posterior so that the true redshift lies within the support of the posterior.
This broadening reflects the increased uncertainty due to the photometric errors.
Finally, the green posterior has had the $u$ band marginalized.
This posterior favors higher redshifts, but still comfortably agrees with the true redshift.
This demonstrates how the $u$ band helps constrain the redshift, and the loss of this information leads to greater uncertainty.

Calculating these posteriors enables direct comparison of true redshift posteriors (consistent with photometric errors and missing bands) with the redshift posteriors estimated by photo-z estimators.
This is important, as modern cosmology analyses are beginning to increasingly rely on full redshift posteriors \citep{mandelbaum2008,newman2022}.
\citePZt showed that popular metrics for evaluating photo-z estimators using ensembles of photo-z posteriors can be misleading, and are not well suited to our science.
PZFlow catalogs with \emph{true} redshift posteriors provide a path forward by enabling the evaluation of photo-z estimators on a per-posterior basis.

@@

% this figure is so high just to move it ahead one page in the document
\begin{figure*}[t]
    \script{photo-z/plot_ensemble_posteriors.py}
    \begin{centering}
        \includegraphics{figures/ensemble_posteriors.pdf}
        \caption{
            The ensemble of posteriors for three example galaxies.
            Flows 1-4 label the individual posteriors produced by each of the flows that make up the ensemble.
            The dashed black line is the mean of these individual posteriors and is the value used by the ensemble.
            The vertical gray line labeled ``Truth'' denotes the true redshift of the galaxy.
            Note these galaxies were specifically chosen for their broad, multimodal posteriors.
            The posteriors of most galaxies are sharp and unimodal.
        }
        \label{fig:ensemble-posteriors}
    \end{centering}
\end{figure*}

In addition to the galaxy magnitudes and redshifts modeled above, we wish to include other galaxy properties in the catalog, such as galaxy size and ellipticity.
In principle, we could have included these variables in the original normalizing flow.
However, we did not want the true redshift posteriors to be conditioned on these variables, as most photo-z estimators only use galaxy photometry.
Therefore, we will build a second flow that models these additional values conditioned on the galaxy redshift and magnitudes.
Note that while we have only chosen to model these additional two properties, any other values you desire can be similarly modeled.

For the latent distribution, we again use a Uniform distribution over the range $[-5, 5]$.
For the bijection, we use
\begin{align}
    \begin{split}
        f =& ~ \text{RQ-RSC} \circ \text{Shift Bounds}. \\
    \end{split}
\end{align}
The RQ-RSC acts on the two dimensional space of size and ellipticity, but also takes the galaxy redshift and magnitudes as inputs (see Figure \ref{fig:conditional-coupling}).
The redshifts and magnitudes are transformed to have zero mean and unit variance before being input to the neural network\footnote{These variables are standard scaled instead of mapped onto the domain [-5, 5], because the neural network that parameterizes the splines has no limit on inputs, unlike the splines themselves, which are limited to the range [-5, 5].} that parameterizes the splines.
Aside from the change in inputs, the RQ-RSC has the same settings as listed for the previous normalizing flow.

After training the flow (see Appendix \ref{app:training-details}), we sample a size and ellipticity for each galaxy in the PZFlow catalog created in the previous section (conditioned on the true magnitudes), and plot the distribution of these features against the distribution in the test set (Figure \ref{fig:conditional-corner}).
Once again, we see the normalizing flow does a good job of emulating the CosmoDC2 galaxy distribution.

The final simulated catalog consists of $10^4$ galaxies, each with a redshift, $ugrizy$ magnitudes including photometric errors, a true photo-z posterior consistent with the photometric errors, a size, and an ellipticity.
This small catalog was generated for visualization purposes, but the normalizing flows can be used to generate catalogs of arbitrarily large size.

>>>>

% Title
\title{Probabilistic Forward Modeling of Galaxy Catalogs with Normalizing Flows}

@@

\begin{abstract}
    Evaluating the accuracy and calibration of the redshift posteriors produced by photometric redshift (photo-z) estimators is vital for enabling precision cosmology and extragalactic astrophysics with modern wide-field photometric surveys.
    Evaluating photo-z posteriors on a per-galaxy basis is difficult, however, as galaxies have a true redshift, but not a true redshift posterior.
    We introduce PZFlow, a Python package for the probabilistic forward modeling of galaxy catalogs with normalizing flows.
    For catalogs simulated with PZFlow, there is a natural notion of ``true'' redshift posteriors that can be used for photo-z validation.
    We use PZFlow to simulate a photometric galaxy catalog where each galaxy has a redshift, noisy photometry, shape information, and a true redshift posterior.
    We also demonstrate the use of an ensemble of normalizing flows for photo-z estimation.
    We discuss how PZFlow will be used to validate the photo-z estimation pipeline of the Dark Energy Science Collaboration (DESC), and the wider applicability of PZFlow for statistical modeling of any tabular data.
\end{abstract}

@@

Photometric redshift (photo-z) estimation is necessary for the study of cosmology and galaxy evolution with the huge galaxy catalogs generated by modern wide-field photometric surveys like the Vera C. Rubin Observatory's Legacy Survey of Space and Time (LSST; \citealt{ivezic2019}).
The colors of galaxies in these surveys are used to estimate posterior distributions over the possible redshift of each galaxy.
These posteriors are often multimodal, reflecting degeneracies in redshift and galaxy spectral type.
Guaranteeing the accuracy and calibration of these posteriors, and understanding their biases, is vital for enabling precision cosmology and studies of galaxy evolution \citep{newman2022,descSRD}.

There have been many studies evaluating the performance of photo-z estimation by comparing estimates to known true redshifts (e.g., \citealt{hildebrandt2010}, \citealt{sanchez2014}, \citealt{graham2018}), however to meet the needs of modern surveys, we must evaluate the accuracy of the full redshift posteriors generated for these galaxies.
In other words, not only must we verify that photo-z estimators assign a high posterior probability to the true redshift, but also that the posteriors account for all the degeneracies inherent in photo-z estimation, and that the distributions are neither too narrow nor too broad.
This kind of validation is difficult, however, because individual galaxies do not have a ``true'' redshift posterior to which you can compare the estimated posterior.
\citePZt studied whether these questions could be answered using ensemble-level comparisons of photo-z posteriors and true redshifts, but found that these metrics could be fooled by a pathological photo-z estimator that ignored galaxy colors altogether.
This raised the need for new methodologies that enable the evaluation of photo-z posterior estimates at the per-galaxy level.

One way to enable posterior validation at the per-galaxy level is to simulate catalogs by sampling galaxies from a probabilistic model for which you have direct access to the underlying probability distribution.
Then, for each galaxy, you can calculate the posterior over redshift with respect to the same model that generated the galaxy.
This provides a ``true'' redshift posterior for each simulated galaxy, to which you can compare the posteriors from photo-z estimators.

Normalizing flows are a generative model of this variety.
Normalizing flows model complex, high-dimensional probability distributions using deep neural networks that learn an invertible mapping between the complicated data distribution and a more simple distribution, known as the latent distribution.
This allows you to sample from and calculate probabilities with respect to the latent distribution, and use the normalizing flow to translate these values back to the space of the original data distribution.
A common choice for the latent distribution is the Normal distribution, hence the name \emph{normalizing} flow.
Unlike other deep generative models like Generative Adversarial Networks (GANs; \citealt{goodfellow2014}) and Variational Autoencoders (VAEs; \citealt{kingma2014}), normalizing flows provide a deterministic mapping between the data and latent spaces.
It is this feature of normalizing flows that enables us to provide an exact answer to the question ``under the generative model, what is the posterior distribution for redshift given the simulated photometry''.
This allows us to define a ``true'' redshift posterior for each galaxy in our simulated catalog, making normalizing flows a very powerful tool for validating photo-z inference.

In this paper, we introduce PZFlow, a normalizing flow package for Python that is designed to facilitate forward modeling galaxy catalogs with true posteriors for galaxy properties.
In addition, we demonstrate PZFlow as a photo-z estimator.
With relatively little tuning required by the user, PZFlow can provide a generative model for any tabular data, including continuous and discrete variables, and variables with Euclidean or periodic topology (e.g. the celestial sphere).
While calculating posteriors, PZFlow can convolve error distributions and marginalize over missing values.

In Section \ref{sec:nf} we provide the background on normalizing flows required to understand PZFlow, which we describe in Section \ref{sec:pzflow}.
In Section \ref{sec:fwd-model}, we demonstrate using PZFlow to simulate a galaxy catalog where each object has a redshift, photometry in the six LSST bands, a true photo-z posterior, a size, and an ellipticity.
In Section \ref{sec:photo-z}, we demonstrate using PZFlow as a density estimator by estimating photo-z's for our simulated catalog.
We conclude in Section \ref{sec:conclusion}.

@@

If we further assume that $f$ is invertible, we can sample from the data distribution by applying $f^{-1}$ to samples from the latent distribution\footnote{Here, $\sim$ means ``is drawn from.''}:
\begin{align}
    x = f^{-1}(u) \quad \text{where} \quad u \sim \pu.
\end{align}

@@

A bijection is an invertible map between two sets.
In a normalizing flow, the bijection maps the data distribution onto the latent distribution for probability calculation, and the inverse of the bijection maps samples from the latent distribution back to the data distribution.
The bijection of a normalizing flow must be powerful enough to model complex relationships in data, while remaining invertible and simultaneously possessing an efficiently computable Jacobian determinant.
This latter constraint is the primary difficulty in designing a normalizing flow.
The most popular strategy for achieving these requirements is to exploit the fact that a composition of bijections is also bijective.
By chaining together multiple less-expressive bijections whose Jacobians are efficiently computable, a composite bijections can be constructed that meets our requirements:
\begin{align}
    f &= \dots \circ f_3 \circ f_2 \circ f_1.
\end{align}
The overall Jacobian determinant can be efficiently calculated using the chain rule.

@@

In general, as long as sampling and density evaluation are tractable, one can use any parameterization of the latent distribution that matches some desired structure in the data and learn the distribution parameters during training.
We give this generalization the name \emph{latent-adaptive flows} (LAFs; inspired by the Tail Adaptive Flows of \citealt{jaini2020}).
Learnable latent distributions can improve training loss, but require more care in training.

@@

PZFlow is a Python package for building normalizing flows, with a focus on features useful for forward modeling and density estimation for tabular data.
Data is handled in Pandas DataFrames \citep{pandas}, while the normalizing flows are implemented in Jax \citep{jax}, which allows for efficient, parallelizable, GPU-enabled calculations for very large data sets.
The code is easily installable from the Python Package Index\footnote{\url{https://pypi.org/project/pzflow/}} (PyPI) and is hosted on Github\footnote{\url{https://github.com/jfcrenshaw/pzflow}}.
The documentation\footnote{\url{https://jfcrenshaw.github.io/pzflow/}} includes tutorial notebooks demonstrating the features mentioned in this paper on different example problems.

The rest of this papaper will focus on forward modeling a photometric galaxy catalog, and photo-z inference.
Section \ref{sec:galaxy-catalog} uses PZFlow to forward model a galaxy catalog, including photometry, spectroscopic redshifts (spec-z's), \emph{true} photo-z posteriors, ellipticities, and sizes.
Section \ref{sec:photo-z} uses PZFlow for photo-z estimation, demonstrating the power of PZFlow as a density estimator, including numerous useful features for photo-z estimation.

In addition to the examples in this paper, PZFlow has already being used in various other DESC projects:
\begin{itemize}
    \item \citet{malz2021} used PZFlow to build a metric for observing strategy optimization based on information theory;
    \item \citet{stylianou2022} used PZFlow to forward model galaxy data with true redshift posteriors in order to evaluate the impact of survey incompleteness and spec-z errors on photo-z estimation;
    \item \citet{lokken2022} used PZFlow to smooth high-redshift artifacts in simulations of host galaxies for supernovae and other transients.
\end{itemize}

@@

In this section, we use PZFlow to forward model a photometric galaxy catalog for the Vera Rubin Observatory's Legacy Survey of Space and Time (LSST; \citealt{ivezic2019}).
The advantage of using a catalog generated from a normalizing flow is that we have direct access to the probability distribution from which the data is drawn, enabling us to calculate true values for derived statistical products, such as the \emph{true} photo-z redshift posterior for each galaxy.

@@

\begin{figure*}[t]
    \script{forward_model/plot_main_galaxy_corner.py}
    \begin{centering}
        \includegraphics{figures/main_galaxy_corner.pdf}
        \caption{
            Distribution of true redshifts and noisy photometry from the CosmoDC2 test set, compared to a sample drawn from the distribution learned by PZFlow.
            The close overlap of every pair-wise distribution demonstrates that PZFlow has learned the distribution in CosmoDC2 with high fidelity.
        }
        \label{fig:main-corner}
    \end{centering}
\end{figure*}

\begin{figure*}[t]
    \script{forward_model/plot_smooth_color_distribution.py}
    \begin{centering}
        \includegraphics{figures/smooth_color_distribution.pdf}
        \caption{
            Comparing the $r-i$ vs redshift distribution for galaxy samples from CosmoDC2 without photometric noise (left), CosmoDC2 with photometric noise (middle), and from the normalizing flow (right).
            High-redshift galaxies in CosmoDC2 lie along discrete tracks in redshift-color space.
            Adding photometric noise ameliorates but does not totally remove the problem.
            PZFlow produces a catalog with a smooth redshift-color distribution.
        }
        \label{fig:smooth-color-dist}
    \end{centering}
\end{figure*}

To create a generative model of galaxy redshifts and photometry, we use the true redshifts and $ugrizy$ magnitudes from the CosmoDC2 simulation \citep{dc2, cosmodc2} of the LSST Dark Energy Science Collaboration (DESC).
We add photometric errors to the true $ugrizy$ magnitudes using the 10-year-depth LSST extended-source error model of our PhotErr package (see Appendix \ref{app:error-model}), and selected galaxies with a signal-to-noise ratio (SNR) greater than 5 in the $i$ band.
Of these, we randomly selected $10^6$ galaxies and split them into training and test sets consisting of 80\% and 20\% of the galaxies, respectively.

@@

After training the flow (see Appendix \ref{app:training-details}), we assess the results by drawing $10^4$ galaxies from the trained flow, and plotting their distribution against  $10^4$ galaxies from the test set (Figure \ref{fig:main-corner}).
We see the normalizing flow has done an excellent job of reproducing the distribution of galaxies in CosmoDC2, without any unusual artifacts or outliers.
In addition, Figure \ref{fig:smooth-color-dist} compares the distribution of galaxy $r-i$ vs redshift.
The CosmoDC2 simulation is known to exhibit discrete tracks in this space at high redshift, due to the way galaxies were assigned to a discrete set of SED templates during simulation.
These tracks are easily visible in the left panel.
Adding photometric errors largely smooths the distribution, but close inspection reveals that discrete tracks are still present in the data, especially at $z > 1$.
The right panel shows the distribution produced by PZFlow, which has a smooth color distribution, even at high redshift.
We note that these results were obtained without any extensive hyperparameter search, and that very similar (slightly worse results) are obtained without the \texttt{ColorTransform} bijection, demonstrating the flexibility of the method to adapt to unseen data sets.

With this normalizing flow, we have an efficient, probabilistic CosmoDC2 emulator that models a smooth color-redshift distribution up to redshift 3.
We generate a catalog by sampling $10^4$ galaxies from the flow, each with noisy photometry and a true redshift.
Importantly, since we have access to the probability distribution from which the galaxies were generated, we can calculate \emph{true} redshift posteriors for each galaxy.
This is the subject of the next section.

@@

\begin{figure}[t]
    \script{forward_model/plot_posteriors.py}
    \begin{centering}
        \includegraphics{figures/posteriors.pdf}
        \caption{
            Example of redshift posteriors for a galaxy simulated with PZFlow.
            The true redshift of the galaxy is marked by the vertical dashed red line, and the true redshift posterior for the galaxy is drawn in black.
            Calculating the posterior while marginalizing over the $u$ band magnitude yields the posterior in blue.
            Note that the $u$ band marginalization is only approximate, but increasing the resolution of the grid of $u$ band values causes the resultant posterior to converge.
        }
        \label{fig:posteriors}
    \end{centering}
\end{figure}

@@

Since we have direct access to the probability distribution from which the photometry and redshifts are drawn, we can analytically calculate the true redshift posterior for each galaxy: $p(z|\mathbf{\hat{m}})$ where $\mathbf{\hat{m}}$ is the vector of noisy galaxy magnitudes.
We note this is not an estimate, like would be returned by a photo-z estimator, but rather the truth, obtained from the model that generated the photometry and redshifts in the first place.

When calculating these posteriors, for each galaxy we can also marginalize over the magnitudes in any missing bands, $n$:
\begin{align}
    p(z|\mathbf{\hat{m}}) = \frac{1}{p(\mathbf{\hat{m}})} \int p(z, n, \mathbf{\hat{m}}) dn,
\end{align}
which can be calculated by evaluating $p(z, n, \mathbf{\hat{m}})$ on a grid of $z$ and possible values of $n$, summing over $n$ to yield $p(z, \mathbf{\hat{m}})$, and normalizing with respect to redshift to yield $p(z|\mathbf{\hat{m}})$.
You may wish to marginalize over all values of $n$ if the galaxy was not observed in that band.
This may occur, for example, when simulated a joint Euclid-LSST catalog \citep{euclid}, as not all galaxies will have photometry from both.
You may also wish to marginalize over all values beyond a limiting magnitude to simulate a galaxy that was observed but not detected in the corresponding band.
This might occur, for example, in the low wavelength bands of Lyman-dropout galaxies observed by LSST.

Note that this marginalization is only approximate, is weakens our ability to refer to these as ``true'' redshift posteriors.
However, increasing the resolution of the grid used for $u$ band marginalization causes the resultant posteriors to converge.
Thus, we believe it is still appropriate to treat these marginalized posteriors as the truth for the purpose of photo-z validation.

Redshift posteriors for an example galaxy are displayed in Figure \ref{fig:posteriors}.
The black posterior is calculated using the full set of galaxy magnitudes.
The true redshift, marked with by the vertical red line, nearly coincides with the mode of this posterior.
The blue posterior has been calculated while marginalizing over the $u$ band.
Throwing away the information in the $u$ band slightly broadens the posterior and shifts it toward higher redshifts.

Calculating these posteriors enables direct comparison of true redshift posteriors (consistent with photometric errors and missing bands) with the redshift posteriors estimated by photo-z estimators.
This is important, as modern cosmology analyses are beginning to increasingly rely on full redshift posteriors \citep{mandelbaum2008,newman2022}.
\citePZt showed that popular metrics for evaluating photo-z estimators using ensembles of photo-z posteriors can be misleading, and are not well suited to the needs of precision cosmology.
PZFlow catalogs with true redshift posteriors provide a path forward by enabling the evaluation of photo-z estimators on a per-posterior basis.

@@

% this figure is so high just to move it ahead one page in the document
\begin{figure*}[t]
    \script{photo-z/plot_ensemble_posteriors.py}
    \begin{centering}
        \includegraphics{figures/ensemble_posteriors.pdf}
        \caption{
            The ensemble of posteriors for three example galaxies.
            Flows 1-4 label the individual posteriors produced by each of the flows that make up the ensemble.
            The dashed black line is the mean of these individual posteriors and is the value used by the ensemble.
            The vertical gray line labeled ``Truth'' denotes the true redshift of the galaxy.
            Averaging the posteriors from each flow in the ensemble approximately marginalizes over the neural network parameters, and smooths over the small-scale variations found in the posterior from any individual flow.
        }
        \label{fig:ensemble-posteriors}
    \end{centering}
\end{figure*}

In addition to the galaxy magnitudes and redshifts modeled above, we wish to include other galaxy properties in the catalog, such as galaxy size and ellipticity.
In principle, we could have included these variables in the original normalizing flow.
However, we did not want the true redshift posteriors to be conditioned on these variables, as most photo-z estimators only use galaxy photometry.
Therefore, we will build a second flow that models these additional values conditioned on the galaxy redshift and magnitudes.
In other words, we are modeling the full joint distribution via the decomposition
\begin{align}
    p(z, \mathbf{\hat{m}}, s, e) = p(z, \mathbf{\hat{m}}) \cdot p(s, e | z, \mathbf{\hat{m}}),
    \label{eq:chain-rule}
\end{align}
where $s$ is the size (the half-light radius in arcseconds) and $e$ is the ellipticity.
The first distribution on the right hand side of Equation \ref{eq:chain-rule} is modeled by our original flow, and the second distribution will be modeled using the new conditional flow.
Note that while we have only chosen to model these additional two properties, any other values you desire can be similarly modeled.

For the latent distribution, we again use a Uniform distribution over the range $[-5, 5]$.
For the bijection, we use
\begin{align}
    \begin{split}
        f =& ~ \text{RQ-RSC} \circ \text{Shift Bounds}. \\
    \end{split}
\end{align}
The RQ-RSC acts on the two dimensional space of size and ellipticity, but also takes the galaxy redshift and magnitudes as inputs (see the conditional variables in green in Figure \ref{fig:conditional-coupling}).
The redshifts and magnitudes are transformed to have zero mean and unit variance before being input to the neural network\footnote{These variables are standard scaled instead of mapped onto the domain [-5, 5], because the neural network that parameterizes the splines has no limit on inputs, unlike the splines themselves, which are limited to the range [-5, 5].} that parameterizes the splines.
Aside from the change in inputs, the RQ-RSC has the same settings as listed for the previous normalizing flow.

After training the flow (see Appendix \ref{app:training-details}), we sample a size and ellipticity for each galaxy in the PZFlow catalog created in the previous section (conditioned on the redshift and noisy magnitudes), and plot the distribution of these features against the distribution in the test set (Figure \ref{fig:conditional-corner}).
Once again, we see the normalizing flow does a good job of emulating the CosmoDC2 galaxy distribution.
